---
title: "PEF"
author: "Johannes Rajala"
date: "`r Sys.Date()`"
output: github_document
---

### PEF-values and asthma

In this document the PEF-values of a person will be analyzed in order to determine whether the person might have asthma or not. The data consists of PEF-values measured in the morning and in the evening. Six measurements are taken each time, three before and three after taking asthma medication. The data also contains the day of the week.

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(aaltobda)
library(ggplot2)
library(ggridges)
library(bayestools)
library(gridExtra)
library(dplyr)
library(rstan)
library(loo)
library(lme4)
library(brms)
cores = parallel::detectCores() / 8
options(mc.cores = cores)
```

Asthma: if the relative effect of medication is over 15$\%$ with at least 3/14 probability OR the relative effect of time of day is over 20$\%$ with at least 3/14 probability, evidence for asthma is present. A reference value for the mean PEF-value for the person in question is 442 liters per minute.

```{r, echo=FALSE}
pefdata <- read.csv("~/Desktop/R_projektit/PEF/pefdata.csv")
```

Let us define a Bayesian linear model for the PEF measurements and draw samples from the posterior distribution.

```{stan, output.var="simple", echo=FALSE, message=FALSE, warning=FALSE, include=FALSE, results="hide"}
data {
  int<lower=0> N; // number of observations
  array[N] real<lower=0> y; // PEF measurements
  array[N] int<lower=0,upper=1> x; // medication indicator (0=before, 1=after)
  array[N] int<lower=0,upper=1> t; // time of day indicator (0=morning, 1=evening)
  array[N] int<lower=1,upper=7> d; // cat presence indicator (1=cat present, 0=cat absent)
}

parameters {
  real alpha; // intercept
  real beta1; // effect of medication
  real beta2; // effect of time of day
  real beta3; // effect of day of the week
  real<lower=0> sigma; // standard deviation
}

transformed parameters {
  vector[N] mu;
  for (i in 1:N)
    mu[i] = alpha + beta1 * x[i] + beta2 * t[i] + beta3 * d[i];
}

model {
  // priors
  alpha ~ normal(300, 100); // weakly informative prior for the intercept
  beta1 ~ normal(0, 100); // weakly informative prior for the effect of medication
  beta2 ~ normal(0, 100); // weakly informative prior for the effect of time of day
  beta3 ~ normal(0, 100); // weakly informative prior for the effect of day of the week
  sigma ~ normal(0, 100); // half-normal due to constraint for the standard deviation

  // likelihood
  y ~ normal(mu, sigma);
}
```

```{r, echo=FALSE, results="hide"}
stan_data <- list(
  y = pefdata$Measurement,
  x = pefdata$Medication,
  t = as.integer(pefdata$Time_of_day == "Evening"),
  d = pefdata$Day,
  N = nrow(pefdata)
)

# Draw samples
model_simple <- rstan::sampling(simple, stan_data, chains = cores, iter = 100000)
draws_simple <- rstan::extract(model_simple)
```

Lets plot the posterior distribution of $\alpha$ (the mean value) with a 95$\%$ credible interval, with the reference value of 442.

```{r, echo=FALSE}
p = mean(draws_simple$alpha > 442)
ci = quantile(draws_simple$alpha, c(0.025, 0.975))

# plot alpha and its ci
ggplot(as.data.frame(draws_simple)) +
  geom_density(aes(x=alpha), fill="blue", alpha=0.5) +
  geom_vline(xintercept=442, linetype="dashed", color="purple") +
  geom_vline(xintercept=ci[1], linetype="dashed", color="red") +
  geom_vline(xintercept=ci[2], linetype="dashed", color="red") +
  annotate("text", x=450, y=0.1, label=paste("P(alpha > 442) = ", round(p, 4))) +
  theme_minimal()
```

The credible interval covers the reference value of 442, and the probability that $\alpha > 442$ is approximately 0.19$\%$.

```{r, echo=FALSE}
# 95$\%$ credible intervals for medication on time of day
medication_effect <- draws_simple$beta1
time_of_day_effect <- draws_simple$beta2
day_of_week_effect <- draws_simple$beta3

# Lets calculate the relative effects of medication and time of day

# alpha plus time 
alpha_time <- draws_simple$alpha + draws_simple$beta2
# alpha + medication
alpha_med <- draws_simple$alpha + draws_simple$beta1

# alpha_time / alpha 
alpha_time_alpha <- alpha_time / draws_simple$alpha
# alpha_med / alpha
alpha_med_alpha <- alpha_med / draws_simple$alpha

# Probability that the relative effect of medication is over 15$%
p_med <- mean(alpha_med_alpha < 1.15)

# Probability that the relative effect of time of day is over 20$%
p_time <- mean(alpha_time_alpha < 1.20)

# Plot the relative effects
ggplot() +
  geom_density(aes(x=alpha_time_alpha), fill="blue", alpha=0.5) +
  geom_density(aes(x=alpha_med_alpha), fill="red", alpha=0.5) +
  geom_vline(xintercept=1.15, linetype="dashed", color="red") +
  geom_vline(xintercept=1.20, linetype="dashed", color="blue") +
  annotate("text", x=1.13, y=50, label=paste("P(relative effect of time < 1.20) = ", round(p_time, 4))) +
  annotate("text", x=1.13, y=45, label=paste("P(relative effect of meds < 1.15) = ", round(p_med, 4))) +
  theme_minimal()
```

Both the relative effect of medication and time of day are less than the thresholds of 15$\%$ and 20$\%$, respectively, with probabilities of 1 and 1.

