---
title: "Analysis of PEF-values with bayesian methods"
author: "Johannes Rajala"
date: "`r Sys.Date()`"
output: github_document
---

### 1. Introduction

In this document, we will analyze the peak expiratory flow (PEF) values of a person in order to gather evidence whether the person might have asthma or not. These values alone are not enough to make a diagnosis, but they can be used as a part of the diagnostic process.

The data consists of PEF-values measured in the morning and in the evening. Six measurements are taken each time, three before and three after taking asthma medication. The data was gathered over a period of seven days. 


```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(ggplot2)
library(ggridges)
library(bayestools)
library(gridExtra)
library(dplyr)
library(rstan)
library(loo)
library(brms)
cores = parallel::detectCores() / 2
options(mc.cores = cores)
```

We consider the following to be evidence for asthma: 

1. The relative effect of medication (variable name $\textit{treatment}$) is over 15$\%$ with at least 3/14 probability.

2. The relative effect of time of day (variable name $\textit{circadian}$) is over 20$\%$ with at least 3/14 probability.

And last, the reference value for the mean PEF-value for the person in question (variable name $\alpha$) is 442.


```{r, echo=FALSE}
# set current project folder as working directory
setwd(here::here())
pefdata <- read.csv("./pefdata.csv")
# encode morning and evening as 0 and 1
pefdata$Time_of_day <- ifelse(pefdata$Time_of_day == "Morning", 0, 1)
```

### 2. Model

The model we will use is a Bayesian linear model. We will model the PEF measurements as a linear combination of the medication effect, the time of day effect, and the day of the week effect. The model is defined as follows:

$y \sim N(\alpha + \beta_1  x + \beta_2  t + \beta_3  d, \sigma)$,

where: $\alpha$ is the intercept, $x$ is the medication indicator, $t$ is the time of day indicator, $d$ is the day of the week, and $\sigma$ is the standard deviation.

We will use weakly informative normal prior distributions for the parameters. The intercept $\alpha$ is given a normal prior with mean 400 and standard deviation 100. The effects of medication, time of day, and day of the week are given normal priors with mean 0 and standard deviation 100. The standard deviation $\sigma$ is given a half-normal prior with standard deviation 100.

### 3. Analysis

#### 3.1. Model in Stan

Let us define a Bayesian linear mode for the PEF measurements in Stan and draw samples from the posterior distribution.

```{stan, output.var="simple", echo=TRUE, results="hide"}
data {
  int<lower=0> N; // number of observations
  array[N] real<lower=0> y; // PEF measurements
  array[N] int<lower=0,upper=1> x; // medication indicator (0=before, 1=after)
  array[N] int<lower=0,upper=1> t; // time of day indicator (0=morning, 1=evening)
  array[N] int<lower=1,upper=7> d; // day of the follow-up
}

parameters {
  real alpha; // intercept
  real treatment; // effect of medication
  real circadian; // effect of time of day
  real followup; // effect of day of the week
  real<lower=0> sigma; // standard deviation
}

transformed parameters {
  vector[N] mu;
  for (n in 1:N) {
    mu[n] = alpha + treatment * x[n] + circadian * t[n] + followup * d[n];
  }
}

model {
  // priors
  alpha ~ normal(400, 100); // weakly informative prior for the intercept
  treatment ~ normal(0, 100); // weakly informative prior for the effect of medication
  circadian ~ normal(0, 100); // weakly informative prior for the effect of time of day
  followup ~ normal(0, 100); // weakly informative prior for the effect of day of the week
  sigma ~ normal(0, 100); // half-normal due to constraint for the standard deviation

  // likelihood
  y ~ normal(mu, sigma);
}

generated quantities {
  vector[N] log_lik;
  for (n in 1:N) {
    log_lik[n] = normal_lpdf(y[n] | mu[n], sigma);
  }
}
```

```{r, echo=FALSE, results="hide"}
stan_data <- list(
  y = pefdata$Measurement,
  x = pefdata$Medication,
  t = pefdata$Time_of_day,
  d = pefdata$Day,
  N = nrow(pefdata)
)

# Draw samples
model_simple <- rstan::sampling(simple, stan_data, chains = cores, iter = 10000)
draws_simple <- rstan::extract(model_simple)
```

#### 3.2. Diagnostics

Let's check diagnostics.

```{r, echo=FALSE}
monitor = rstan::monitor(model_simple, print = FALSE)
monitor[seq(1,5), c(1,2,3,10)]
```

$\hat R$ diagnostic values are below 1.01, which indicates that the chains have converged. Let's move onto the analysis.

#### 3.3. Results

Lets plot the posterior distribution of $\alpha$ (the mean value) with a 95$\%$ credible interval, with the reference value of 442.

```{r, echo=FALSE}
p = mean(draws_simple$alpha > 442)
ci = quantile(draws_simple$alpha, c(0.025, 0.975))

# plot alpha and its ci
p1 = ggplot(as.data.frame(draws_simple)) +
  geom_density(aes(x=alpha), fill="blue", alpha=0.5) +
  geom_vline(xintercept=442, linetype="dashed", color="purple") +
  geom_vline(xintercept=ci[1], linetype="dashed", color="red") +
  geom_vline(xintercept=ci[2], linetype="dashed", color="red") +
  annotate("text", x=450, y=0.11, label=paste("P(mean PEF value > 442) = ", round(p, 4))) +
  xlab("Mean PEF value") +
  ggtitle("Posterior distribution of alpha", subtitle = "With 95% credible interval and the reference value") +
  theme_minimal()

p1
```

The probability that $\alpha > 442$ is approximately 0.19$\%$.

```{r, echo=FALSE}
# 95$\%$ credible intervals for medication on time of day
medication_effect <- draws_simple$treatment
time_of_day_effect <- draws_simple$circadian
day_of_week_effect <- draws_simple$followup

# Lets calculate the relative effects of medication and time of day

# alpha plus time 
alpha_time <- draws_simple$alpha + draws_simple$circadian
# alpha + medication
alpha_med <- draws_simple$alpha + draws_simple$treatment

# alpha_time / alpha 
alpha_time_alpha <- alpha_time / draws_simple$alpha
# alpha_med / alpha
alpha_med_alpha <- alpha_med / draws_simple$alpha

# Probability that the relative effect of medication is over 15$%
p_med <- mean(alpha_med_alpha < 1.15)

# Probability that the relative effect of time of day is over 20$%
p_time <- mean(alpha_time_alpha < 1.20)

# Plot the relative effects
p2 = ggplot() +
  geom_density(aes(x=alpha_med_alpha), fill="blue", alpha=0.5) +
  geom_vline(xintercept=1.15, linetype="dashed", color="purple") +
  annotate("text", x=1.1, y=45, label=paste("P(relative effect of medication < 15%) = ", round(p_med, 4))) +
  xlab("Relative effect of medication") +
  ggtitle("Posterior distribution of the relative effect of medication", subtitle = "With 15% threshold") +
  theme_minimal()

p3 = ggplot() +
  geom_density(aes(x=alpha_time_alpha), fill="blue", alpha=0.5) +
  geom_vline(xintercept=1.20, linetype="dashed", color="purple") +
  annotate("text", x=1.1, y=45, label=paste("P(relative effect of time of day < 20%) = ", round(p_time, 4))) +
  xlab("Relative effect of time of day") +
  ggtitle("Posterior distribution of the relative effect of time of day", subtitle = "With 20% threshold") +
  theme_minimal()

p2

p3

```

Both the relative effect of medication and time of day are less than the thresholds of 15$\%$ and 20$\%$, respectively, with probabilities of 1 and 1.

No strong evidence for the person having asthma is present.

